Decentralized state machine using Nakamoto (probabilistic) consensus\\\medskip Research Project 2022

### 2022-11-01 article/article.tex
more article

\documentclass[11pt, twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{fdsymbol}
\usepackage{textgreek}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{url}
\usepackage{array}
\usepackage{csquotes}
\usepackage{caption}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}

@@

\tableofcontents
\section{Traditional state machine replication}

@@

To tackle this issue, BFT algorithms were designed \cite{pbft}. They include a layer of authentification, but are not necessarily much slower. However, there is a maximum number of malicious nodes that the algorithm can handle safely \cite{partialSynchrony}. 
For example, assuming partial synchrony of the nodes, consensus protocols can handle t crash failures with 2t+1 nodes but t byzantine failures with no less than 3t+1 nodes.

@@

The Naka consensus
The Naka consensus
The Naka consensus
The Naka consensus
The Naka consensus
The Naka consensus
The Naka consensus

@@

Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine

@@

Markov
Markov
Markov
Markov
Markov
Markov
Markov
Markov
Markov
Markov

@@

\bibliographystyle{alpha}
\bibliography{biblio}

\end{document}

>>>>

\documentclass[11pt, twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{fdsymbol}
\usepackage{textgreek}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{nameref}
\usepackage{url}
\usepackage{array}
\usepackage{csquotes}
\usepackage{caption}
\usepackage{algpseudocode}
\usepackage[toc,page]{appendix}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
\providecommand{\myceil}[1]{\left \lceil #1 \right \rceil }
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}

@@

\tableofcontents

\section{Traditional state machine replication}

@@

To tackle this issue, BFT algorithms were designed (such as \cite{pbft}). They include a layer of authentification, but are not necessarily much slower. However, there is a maximum number of malicious nodes that the algorithm can handle safely \cite{partialSynchrony}.
For example, assuming partial synchrony of the nodes, consensus protocols can handle t crash failures with 2t+1 nodes but t byzantine failures with no less than 3t+1 nodes.

@@

Nakamoto consensus was introduced with the bitcoin blockchain \cite{bitcoin}. Nakamoto consensus is probabilistic, which means that there is a probability ε that consensus is not reached.
Of course, ε can be so low that crucial systems can be built relying Nakamoto consensus. Nakamoto consensus is open : nodes don't have to register or get an authorisation to enter the network.
Because nodes don't have to know every other node in the network and be able to communicate with everyone, Nakamoto is supposed to allow for greater scale in the network.

Nakamoto consensus definetly shaked the consensus space in computer science with cryptocurrencies. In a different context, can it be useful as well ?

@@

To implement Nakamoto consensus, I chose the Snowball consensus algorithm \cite{snowprotocol}, which underlies the Avalanche blockchain.

The goal of snowball is to reach binary consensus starting from a network of nodes.

I adapted the protocol to fit the needs of a decentralized state machine, because snowball is described in the context of building a payment system (because it is the basis of Avalanche, which is that payment system).
Snowball is built by iterations, but I will only focus on Snowball, the last iteration, because it is the most interesting and the one I tested the performances in the end. Slush and Snowflake are interesting to understand
where we are coming from, but no more.

I adapted Snowball in the following ways : Each node, in addition to having all the snowball consensus related data (not much), has a version of the record. All nodes starts with a empty record.
Because it is the requirement for a decentralized state machine, each node can be queried to add a log to the record. Each node can also be queried to get one log in the record.

The snowball protocol is triggered when the client asks to add a new log to an index that has not been touched yet. Indeed, the system works as if each index in the record corresponds to an independant snowball run.

In snowball, each node has a counter of "confidence" which can tip towards one way or another. In my implementation, there is no binary consensus because the only value the system has is the log request.
When receiving an add-log request, the node follows this sequence :
\begin{algorithmic}
    \State from request (client or other node):
    \State $logToAdd$ (string)
    \State $logPosition$ (int)

    \State \\from node memory :
    \State $record$ (map[int]string)
    \State $cnt$ (map[int]int)
    \\
    \If{$record[logPosition]$ exists}
    \State reply $record[logPosition] == logToAdd$
    \Else
    \State $record[logPosition] \gets logToAdd$
    \State reply $True$
    \State initiateRequest($logToAdd$, $logPosition$)
    \EndIf

\end{algorithmic}

initiateRequest corresponds to the following sequence :

\begin{algorithmic}
    \While{$True$}
    \State $success \gets initiateQuery(req)$
    \State $initiateQuery$ picks a sample in the network and queries it
    \If{$success$}
    \State $cnt[logPosition]++$
    \Else
    \If{$cnt[logPosition] == 1$}
    \State delete($record$, $logPosition$)
    \State $cnt[logPosition] \gets 0$
    \Else
    \State $cnt[logPosition]--$
    \EndIf
    \EndIf
    \If{$cnt[logPosition] \geqslant CounterThreshold$}
    \State $record[logPosition] \gets logToAdd$
    \State return
    \EndIf
    \EndWhile
\end{algorithmic}

3 parameters are used in my implementation :
\begin{itemize}
    \item sampleSize (k) : the number of nodes that are queried each round
    \item majorityThreshold (α) : the number of nodes that have to reply successfully to consider that the query is successfull.
          (with \begin{equation*} \alpha \geq \floor{k/2}  \end{equation*})
    \item counterThreshold (β) : the threshold of successfull requests above which the node stops initiating requests and only replies with its content. (for one position)
\end{itemize}

@@

In order to compare the performances of classical consensus and Nakamoto consensus, we need to be able to know the probability of not reaching consensus ε corresponding to the set 
of values (n, k, β). (α is not included because we assume no byzantine nodes in this part).

The evolution of the system of nodes, for one position (one snowball run) is a Markov process. To compute to probability of not reaching consensus, we need to identify the possible states in the 
Markov process, the possible transitions and the transitions probabilities.

\paragraph{β = 1}
With β = 1, the state is (c, reqs) with :
\begin{itemize}
    \item c = Number of colored nodes ($c \leq n$)
    \item reqs = Number of groups of requests which are queried for next round. (They all come from neutral nodes)
\end{itemize}
If the state converges to $c = n$, consensus is reached.
If the state converges to $reqs = 0$, with $c \neq 0$, consensus is not reached when the snowball run is done.

Then, the probability of reaching a certain state (1) is the sum of the probability of a different state (2) times the probability of going from state 2 to state 1.
That is true if and only if states are transient, which is the case here because c can only grow and with c constant, reqs can only diminish.

Therefore, the probability of being at state (c, reqs) is (the term in the sum is considered 0 if it does not make sense, which is the case for some t) :
The sum represents a step of the resolution of one group of requests (t represent the number of neutral nodes touched)
\begin{equation*}
    P(c, reqs) = \sum_{t=0}^{k+1} {n-c+t \choose t}{c-t-1 \choose k-t}P(c-t, reqs+1-t)
\end{equation*}

\paragraph{β = 2}
With β = 2, the state is (c, s, rc, rs), with:
\begin{itemize}
    \item c = Number of colored nodes (cnt = 2)
    \item s = Number of semi-colored nodes (cnt = 1)
    \item rc = Number of groups of requests from colored nodes
    \item rs = Number of groups of requests from semi-colored nodes
\end{itemize}

We must separate requests in those two groups because a request cannot be directed to the node initiating it. 
Therefore the probabilities of hitting a colored node is smaller if the request comes from a colored node than from a semi-colored node.

The term in the sums is considered 0 if it does not make sense.
First sum represents a step of the resolution of one group of requests, coming from a semi-colored node, with tc (ts) the number of colored (semi colored) nodes in the sample.
Second sum represents a step of the resolution of one group of requests, coming from a colored node. 
To do the calculation, we assume that requests from semi-colored nodes resolve faster.
\begin{multline*}
    P(c, s, rc, rs) = \\ \sum_{tv=0}^{k+1}\sum_{ts=0}^{k+1} {n-c-s+tv \choose tv}{s-tv-1+ts \choose ts}{c-ts \choose k-tv-ts}P(c-ts, s-tv+ts, rc-ts, rs+1-tv)
    \\ + \sum_{ts=0}^{k+1} {n-c-s+rs \choose tv}{s-rs+ts \choose ts}{c-ts-1 \choose k-rs-ts}P(c-ts, s-rs+ts, rc+1-ts, 0)
\end{multline*}
\\
With these formulas, we can compute the probability of not reaching consensus : view Appendices

@@

\onecolumn
\bibliographystyle{alpha}
\bibliography{biblio}

\begin{appendices}
    \label{Appendices}
    \begin{figure*}
        Probability of not reaching consensus (β = 1) : ε 
        \centering
        \includegraphics[width=16.5cm]{images/smallNoConsensusBetaOne.png}
    \end{figure*}
    \begin{figure*}
        Probability of not reaching consensus (β = 1) : -log(ε) 
        \centering
        \includegraphics[width=16.5cm]{images/smalllogNoConsensusBetaOne.png}
    \end{figure*}
    \begin{figure*}
        Probability of not reaching consensus (β = 2) : ε 
        \centering
        \includegraphics[width=16.5cm]{images/smallnoconsensus.png}
    \end{figure*}
    \begin{figure*}
        Probability of not reaching consensus (β = 2) : -log(ε) 
        \centering
        \includegraphics[width=16.5cm]{images/smalllognoconsensus.png}
    \end{figure*}
\end{appendices}
\end{document}
